### 2016-10-16 dissertation/normal.tex
dissertation: part normal update 6: finish NB multiplier design

Now let us observe the behavior of 5-bit Agnew's SMPO. Initially all DFFs are reset to 0. 
In the first clock cycle,
signal sent to the flip-flop in block $R_0$ denotes function:
$$R_0^{(1)} = a_1b_0$$
It equals to the first term of equation \ref{eqn:5bitSMPO_c0}. In the second clock cycle,
this signal is sent to block $R_1$ through wire $r_0$, and this block also receives data 
from operands (shifted by 1 bit), generating signal $a_u,a_v$ and $b_w$. Concretely,
signal sent to flip-flop in block $R_1$ is:
$$R_1^{(2)} = R_0^{(1)} + (a_0+a_3)b_1 = a_1b_0+(a_0+a_3)b_1$$
which forms first 2 terms of equation \ref{eqn:5bitSMPO_c0}. Similarly, we track the signal 
on $R_2$ in third clock cycle, signal on $R_3$ in fourth clock cycle, finally we can get 
$$R_4^{(5)} = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4$$
which equals to $c_0$ in equation \ref{eqn:5bitSMPO_c0}.
After the fifth clock cycle ends, this signal can be detected on wire $r_0$. It shows that 
the result of $c_0$ is computed after 5 clock cycles and gives on $r_0$.

@@

\subsection{Multiplier not based on $\lambda$-Matrix}

>>>>

Now let us observe the behavior of 5-bit Agnew's SMPO. Initially all DFFs are reset to 0. 
In the first clock cycle,
signal sent to the flip-flop in block $R_0$ denotes function:
$$R_0^{(1)} = a_1b_0$$
It equals to the first term of equation \ref{eqn:5bitSMPO_c0}. In the second clock cycle,
this signal is sent to block $R_1$ through wire $r_0$, and this block also receives data 
from operands (shifted by 1 bit), generating signal $a_u,a_v$ and $b_w$. Concretely,
signal sent to flip-flop in block $R_1$ is:
$$R_1^{(2)} = R_0^{(1)} + (a_0+a_3)b_1 = a_1b_0+(a_0+a_3)b_1$$
which forms first 2 terms of equation \ref{eqn:5bitSMPO_c0}. Similarly, we track the signal 
on $R_2$ in third clock cycle, signal on $R_3$ in fourth clock cycle, finally we can get 
$$R_4^{(5)} = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4$$
which equals to $c_0$ in equation \ref{eqn:5bitSMPO_c0}.
After the fifth clock cycle ends, this signal can be detected on wire $r_0$. It shows that 
the result of $c_0$ is computed after 5 clock cycles and given on $r_0$.

@@

\subsection{Multiplier not based on $\lambda$-Matrix}
Both Massey-Omura multiplier and Agnew's SMPO rely on the implementation of $\lambda$-Matrix,
which means that they will be identical if unrolled to full combinational circuits. 
After Agnew's work of parallelization, researchers proposed more designs of SMPO, 
some of them jump out of the circle and are independent from $\lambda$-Matrix.
One competitive multiplier design of this type is invented by Reyhani-Masoleh and Hasan 
\cite{RHmulti}, which is therefore called RH-SMPO.

% fig:5bitRH

Figure \ref{fig:5bitRH} is a 5-bit RH-SMPO which is functionally equivalent to 5-bit Agnew's SMPO 
in figure \ref{fig:SMPO}. A brief proof is as follows:

\begin{Proof}
First, we define an auxiliary function for $i$-th bit 
\begin{equation}
\label{eqn:aux}
F_i(A,B) = a_ib_i\beta + \sum_{j=1}^v d_{i,j}\beta^{1+2^j}
\end{equation}
where $0\leq i\leq k-1, v = \lfloor k/2\rfloor, 1\leq j \leq v$.
The $d$-layer index $d_{i,j}$ is defined as
\begin{equation}
\label{eqn:auxDC}
d_{i,j} = c_{a,i} c_{b,i} = (a_i+a_{i+j})(b_i+b_{i+j}),~~1\leq j\leq v
\end{equation}
$i+j$ here is the result reduced modulo $k$. Note that there is a special boundary case when
$k$ is an even number ($v = \frac{k}{2}$):
$$d_{i,v} = (a_i+a_{i+v})b_i$$
With the auxiliary function, we can utilize following theorem (proof refers to \cite{RHmulti}):
\begin{Theorem}
Consider three elements $A,B$ and $R$ such that $R=A\times B$ over $\Fkk$. Then,
$$R=(((F_{k-1}^2+F_{k-2})^2+F_{k-3})^2+\cdots+F_1)^2+F_0$$
\end{Theorem}
This form is called inductive sum of squares, and corresponds to the cyclic shifting on 
$R_i$ flip-flops. Concretely, the multiplier behavior is an implementation of 
following algorithm:

\begin{algorithm}[H]
\SetAlgoNoLine

 \KwIn{$A,B\in \Fkk$ given w.r.t. NB $N$}
 \KwOut{$R=A\times B$}
%%%%%%%%%%%%%%%%%%%%
  Initialize $A,B$ and aux var $X$ to 0\;
  \For { ($i=0$;   $i < k$; ++i ) }
  {
	$X \gets X^2+F_{k-1}(A,B)$ \CommentSty{/*use aux-func from Eq.\ref{eqn:aux}*/\;}
	$A\gets A^2,~B\gets B^2$ \CommentSty{/*Right-cyclic shift $A$ and $B$*/\;}
  }
  $R\gets X$
\caption{NB Multiplication Algorithm in RH-SMPO \cite{RHmulti}}\label{alg:RHmulti}
\end{algorithm}

In this algorithm, we use a fixed auxiliary function $F_{k-1}$ inside the loop.
This is because of equation
$$F_{k-l} = F_{k-1}(A^{2^{l-1}},B^{2^{l-1}}),~~1\le l\le k$$
So using fixed $F_{k-1}$ and squaring $A^{2^i}$ every time inside the loop is equivalent to computing 
$F_{k-1},F_{k-2},\dots,F_0$ with fixed operands $A,B$.
\end{Proof}

To better understand the mechanism of RH-SMPO, we will use this 5-bit 
RH-SMPO as an example and introduce the details on how to design it.
\begin{Example}[Designing a 5-bit RH-SMPO]
From equation \ref{eqn:aux} we can deploy AND gates in $d$-layer according to $d_{i,j}$,
and XOR gates in $c$-layer according to equation \ref{eqn:auxDC}. Concretely, as algorithm \ref{alg:RHmulti}
describes, we implement auxiliary function $F_{k-1}$ in the logic:
$$i = k-1 = 4;~~v=\lfloor 5/2 \rfloor = 2$$
\begin{equation}
\label{eqn:5bitRHaux}
F_{4}(A,B) = a_4b_4\beta+\sum_{j=1}^2 d_{4,j}\beta^{1+2^j} = d_0\beta+\sum_{j=1}^2 d_{4,j}\beta^{1+2^j}
\end{equation}
Consider indices $4+1=0\bmod 5,~4+2=1\bmod 5$, write down gates in $c$-layer and $d$-layer (besides $d_0$)
$$c_1 = a_0+a_4,~c_2 = b_0+b_4,~d_1=d_{4,1}= c_1c_2 = (a_4+a_0)(b_4+b_0)$$
$$c_3 = a_1+a_4,~c_4 = b_1+b_4,~d_2=d_{4,2}= c_3c_4 = (a_4+a_1)(b_4+b_1)$$
The difficult part of the whole design is to deploy XOR gates in $e$-layer. 
As the logic layer closest to the outputs $R_i$, $e$-layer actually finishes the implementation of 
$F_{k-1}(A,B)$. But it is not a simple addition; the reason is before bit-wise adding to $X^2$, it is necessary to 
turn the sum to NB form. In other words, theoretically we need $k$ XOR gates in $e$-layer, the output of 
$i$-th gate corresponds to the bit multiplying $\beta^{2^i}$.

In order to obtain information 
indicating interconnections between $d$-layer and $e$-layer, we need to interpret $\beta^{1+2^j}$ 
to NB representation.
There is a concept called {\bf multiplication table} (M-table) which can assist this interpretation. It is defined as 
a $k\times k$ matrix $T$ over $\mathbb F_2$:
\begin{equation}
\label{eqn:multitable}
\begin{bmatrix}
\beta^{1+2^0} \\ \beta^{1+2^1} \\ \beta^{1+2^2} \\ \vdots \\ \beta^{1+2^{k-1}}
\end{bmatrix}
= \beta
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^4 \\ \vdots \\ \beta^{2^{k-1}}
\end{bmatrix}
=
\begin{bmatrix}
T_{0,0}      &   T_{0,1}        & \dots & T_{0,k-1}\\
T_{1,0}    &   T_{1,1}           & \dots & T_{1,k-1}\\
T_{2,0}    &   T_{2,1}           & \dots & T_{2,k-1}\\
\vdots & \vdots              & \ddots     & \vdots \\
T_{k-1,0}    &   T_{k-1,1}           & \dots & T_{k-1,k-1}
\end{bmatrix}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^4 \\ \vdots \\ \beta^{2^{k-1}}
\end{bmatrix}
= {\bf T}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^4 \\ \vdots \\ \beta^{2^{k-1}}
\end{bmatrix}
\end{equation}

It is a known fact that M-table $T$ can be converted from $\lambda$-Matrix $M$:
$$M_{i,j}^{(0)} = T_{j-i,-i}$$
with indices reduced modulo $k$. Thus we can write down the M-table of $\mathbb F_{2^5}$ with current NB $N$:

% fig:multitable

Note that we only use row 1 and row 2 from the M-table since range $1\leq j \leq 2$.
All nonzero entries in these 2 rows corresponds to the interconnections between $d$-layer and 
$e$-layer. For example, row 1 has two nonzero entries at column 0 and column 3, which corresponds to interconnections 
between $d_1$ and $e_0,e_3$. This conclusion comes from row 1 in equation \ref{eqn:multitable}:
\begin{equation*}
\beta\cdot\beta^2 = 
\begin{bmatrix}
1 & 0 & 0 & 1 & 0
\end{bmatrix}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^4 \\ \beta^8 \\ \beta^{16}
\end{bmatrix}
= \beta + \beta^{2^3}
\end{equation*}

Similarly, from row 2 of M-table we derive that $d_2$ has fanouts $e_3,e_4$:
\begin{equation*}
\beta\cdot\beta^{2^2} = 
\begin{bmatrix}
0 & 0 & 0 & 1 & 1
\end{bmatrix}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^4 \\ \beta^8 \\ \beta^{16}
\end{bmatrix}
= \beta^{2^3} + \beta^{2^4}
\end{equation*}

Let us look back at equation \ref{eqn:5bitRHaux}, we already dealt with the latter part.
The first term is always $d_0\beta$, which denotes $d_0$ should always be connected to $e_0(\beta)$.
After gathering all interconnection information, we can translate it to gate-level circuit implementation:
$$e_0 = d_0+d_1,~e_3=d_1+d_2,~e_4=d_2$$

Then the last mission is to implement the output $R_i$ layer. Assume $r_{i-1}$ is the output of 
$R_{i-1}$ in last clock cycle, we can connect using relation 
$$R_i = r_{i-1} + e_i$$
In this example, according to the M-table in figure \ref{fig:multitable}, columns $e_1,e_2$
have only zeros in its intersection with row $d_1,d_2$. Thus gates for $e_1,e_2$ can be omitted.

This finishes the full design procedure for a 5-bit RH-SMPO.
\end{Example}

The area cost of RH-SMPO is even smaller than Agnew's SMPO. XOR gates corresponds to all nonzero entries 
in M-table, which is with the same number of nonzero entries in $\lambda$-Matrix ($C_N$). The number of AND gates 
equals to $v$ plus 1 (for gate $d_0$). When using ONB ($C_N = 2k-1$), the total number of gates 
is $2k+\lfloor \frac{k}{2}\rfloor$.

\subsection{Customized Galois Field for Optimal Normal Basis Multiplier Design}
